Bayesian inference of prehistoric population dynamics from multiple proxies: a case study from the North of the Swiss Alps

### 2022-11-24 manuscript/paper_revision_1.Rmd
added first revisions of the paper to a separate rmd file

  Here, we present a Bayesian hierarchical model that uses Poisson regression and state-space representation to produce absolute estimates of past population size and density. Using the  area North of the main ridge of the Swiss Alps in prehistoric times (6000–1000 BCE) as a case study, we show that combining multiple proxies (site counts, radiocarbon dates, dendrochronological dates, and landscape openness) produces a more robust reconstruction of population dynamics than any single proxy alone. The model's estimates of the credibility of its prediction, and the relative weight it affords to individual proxies through time, give further insights into the relative reliability of the evidence currently available for paleodemographic research. Our prediction of population development of the case study area accords well with the current understanding in the wider literature, but provides a more precise and higher-resolution estimate that is less sensitive to spurious fluctuations in the proxy data than existing approaches, especially the popular summed probability distribution of radiocarbon dates.
  The archaeological record provides several potential proxies of human population dynamics, but individually they are inaccurate, biased, and sparse in their spatial and temporal coverage. Similarly, current methods for estimating past population dynamics are often simplistic: they work on limited spatial scales, tend to rely ona single proxy, and are rarely able to infer population size or density in absolute terms. In contemporary demography, it is becoming increasingly common to use Bayesian statistics to estimate population trends and project them into the future. The Bayesian approach is popular because offers the possibility of combining heterogenous data, and at the same time qualifying the uncertainty and credibility attached to forecasts. These same characteristics make it well-suited to applications to archaeological data in paleodemographic studies.

@@

Prehistorians have long recognised demography as a fundamental force in human cultural evolution [@childe_man_1936]. Despite decades of interest in the population dynamics of prehistoric societies, concrete estimates of population size and density before written records remain elusive. Though the archaeological record provides multiple possible demographic proxies [@muller_tracing_2019], a lack of access to this data and methodological tools for turning it into quantitative estimates has left the conclusions drawn from it vague and superficial [@hassan_demographic_1981]. As a result, 'expert estimates' transferred from ethnographic parallels have often taken the place of direct inference from archaeological evidence [@morris_measure_2013, @turchin_seshat_2015].

@@

In particular, the 'dates as data' technique [@rick_dates_1987], using the frequency of radiocarbon dates as a proxy for population dynamics, has been significantly developed in the last decade [e.g. @shennan_regional_2013] and widely applied to archaeological contexts worldwide [@crema_review_2022]. This approach has contributed greatly to our understanding of prehistoric demography, but is not without its critics [@attenbrow_hiscock_2015, @carleton_groucutt_2020]. While the methodology continues to evolve and address these critiques [@crema_review_2022], it remains subject to fundamental problems common to all approaches relying on a single proxy [@french_manifesto_2021, @schmidt_approaching_2021]. We believe that these problems cannot be overcome by methodological refinements in this area alone. Instead, a Bayesian approach offers a robust, quantitative methodology for inferring prehistoric population dynamics from multiple proxies, including summed radiocarbon dates.

@@

Our case study area north of the Swiss Alps (Figure \@ref(fig:mapswissplateau)) covers about one third of Switzerland’s territory and comprises the partly flat, but largely hilly area between the Jura Mountains and the Alps. It is favourable for settlement and agriculture; the Swiss Plateau between Lake Zurich and Lake Geneva is by far the most densely populated region of the Switzerland today. This serves as our core region of interest because it is here that archaeological data is most abundant and accessible. The region has a very diverse natural landscape: shaped by glaciers during the ice ages, the many lakes and bogs provide excellent preservation conditions for the numerous Neolithic and Bronze Age lakeside settlements, and a rich source for vegetation reconstructions by means of pollen analyses. Thanks to the very active and efficient archaeological research and heritage management there is an abundance of archaeological information, including known sites as well as dendrochronological and ^14^C data.

@@

: (#tab:tableproxies) A incomplete list of possible observation that can be linked to population developments in the past. Proxies used in this study are highlighted.

@@

@@

We include relative dating information obtained from the heritage authorities of the Swiss cantons (Figure \@ref(fig:aoristmap)). These are primarily derived from scattered surface finds, often with a low dating accuracy (only in the range of archaeological periods), incorporated into our model as a typologically-dated, aorist time series. However, it is not dependent radiocarbon dating and thus it avoids the methodological issues of sum calibration. Data from `r nrow(aorist)` sites were included in the aoristic sum.

@@

We assume that the higher the population density in an area, the greater the human influence on the natural environment [@zimmermann2004, @lechterbeck2014]. Evidence of deforestation can therefore provide indications of population dynamics. The full procedure for deriving this proxy from several different pollen diagrams is detailed in a previous publication [@heitz2021]. Here, we use five pollen diagrams from sites mainly in the hinterland of the large Alpine lakes.

@@

Our overall model is broken down into several hierarchically-connected individual elements. The a process model represents the demographic development itself, without already being explicitly parameterised with data. Here we assume that the latent variable 'number of sites' is strongly autocorrelated across different time periods. The number of sites in 3000 BCE is strongly conditioned by the number of sites in 3050 BCE, and so on. The population at time $t$ results from the population at time $t-1$ times a parameter $\lambda$, which represents the population change at this time.

@@

>>>>

  Here, we present a Bayesian hierarchical model that uses Poisson regression and state-space representation to produce absolute estimates of past population size and density. At its core, the statistical model is as follows: if the proxies as a whole have a positive delta, a Poisson draw for the new number of settlements is biased by a positive amount (and similarly for negative deltas). Using the  area North of the main ridge of the Swiss Alps in prehistoric times (6000–1000 BCE) as a case study, we show that combining multiple proxies (site counts, radiocarbon dates, dendrochronological dates, and landscape openness) produces a more robust reconstruction of population dynamics than any single proxy alone. The model's estimates of the credibility of its prediction, and the relative weight it affords to individual proxies through time, give further insights into the relative reliability of the evidence currently available for paleodemographic research. Our prediction of population development of the case study area accords well with the current understanding in the wider literature, but provides a more precise and higher-resolution estimate that is less sensitive to spurious fluctuations in the proxy data than existing approaches, especially the popular summed probability distribution of radiocarbon dates.
  The archaeological record provides several potential proxies of human population dynamics, but individually they are inaccurate, biased, and sparse in their spatial and temporal coverage. Similarly, current methods for estimating past population dynamics are often simplistic: they work on limited spatial scales, tend to rely on a single proxy, and are rarely able to infer population size or density in absolute terms. In contemporary demography, it is becoming increasingly common to use Bayesian statistics to estimate population trends and project them into the future. The Bayesian approach is popular because it offers the possibility of combining heterogenous data, and at the same time qualifying the uncertainty and credibility attached to forecasts. These same characteristics make it well-suited to applications to archaeological data in paleodemographic studies.

@@

Prehistorians have long recognised demography as a fundamental force in human cultural evolution [@childe_man_1936]. Despite decades of interest in the population dynamics of prehistoric societies, concrete estimates of population size and density before written records remain elusive. Though the archaeological record provides multiple possible demographic proxies [@muller_tracing_2019], a lack of access to this data and methodological tools for turning it into quantitative estimates has left the conclusions drawn from it vague and superficial [@hassan_demographic_1981]. As a result, 'expert estimates' transferred from ethnographic parallels have often taken the place of direct inference from archaeological evidence [@morris_measure_2013; @turchin_seshat_2015].

@@

In particular, the 'dates as data' technique [@rick_dates_1987], using the frequency of radiocarbon dates as a proxy for population dynamics, has been significantly developed in the last decade [e.g. @shennan_regional_2013] and widely applied to archaeological contexts worldwide [@crema_review_2022]. This approach has contributed greatly to our understanding of prehistoric demography, but is not without its critics [@attenbrow_hiscock_2015; @carleton_groucutt_2020; @price_end__end_2021]. While the methodology continues to evolve and address these critiques [@crema_review_2022], it remains subject to fundamental problems common to all approaches relying on a single proxy [@french_manifesto_2021; @schmidt_approaching_2021]. We believe that these problems cannot be overcome by methodological refinements in this area alone. Instead, a Bayesian approach offers a robust, quantitative methodology for inferring prehistoric population dynamics from multiple proxies, including summed radiocarbon dates.

In the following article we present our methodology and demonstrate its application with a case study in the area north of the Swiss Alps. We use a type of Bayesian Generalised Linear Model [@kruschke2015] to combine the different proxies. Their assessment (weight and credibility intervals) results from the modelling and the combination of the data itself and is not defined a priori. Furthermore, we pursue the concept of a state space model [@auger-methe_guide_2021], a time series model in which a time series is interpreted as the result of a noisy observation of a stochastic process. In this sense, it is not a deductive model for testing a hypothesis, but a reconstructive (abductive) one, such it is also used in environmental reconstructions [@inkpen_explaining_2009]. In this respect, we hope to convince the readers that such an approach is far better suited to realising reliable reconstructions of past population developments than any of the previous approaches.

@@

Our case study area north of the Swiss Alps (Figure \@ref(fig:mapswissplateau)) covers about one third of Switzerland’s territory and comprises the partly flat, but largely hilly area between the Jura Mountains and the Alps. It is favourable for settlement and agriculture; the Swiss Plateau between Lake Zurich and Lake Geneva is by far the most densely populated region of the Switzerland today. This serves as our core region of interest because it is here that archaeological data is most abundant and accessible. The region has a very diverse natural landscape: shaped by glaciers during the ice ages, the many lakes and bogs provide excellent preservation conditions for the numerous Neolithic and Bronze Age lakeside settlements, and a rich source for vegetation reconstructions by means of pollen analyses. Thanks to very active and efficient archaeological research and heritage management there is an abundance of archaeological information, including known sites as well as dendrochronological and ^14^C data.

@@

: (#tab:tableproxies) An incomplete list of possible observation that can be linked to population developments in the past. Proxies used in this study are highlighted.

For all proxies used, it is a common assumption that they are positively correlated with past population levels. In some cases we have reprocessed them in this respect in order to emphasise this relationship even more strongly in our opinion (example: pollen data). This is of course a strong assumption, but we believe it is justified, as this has been widely accepted in the scientific literature, and these proxies are used accordingly (each in isolation). Our ambition with this article is to present an abductive model to help improve the current practice of proxy use. Our ambition is not to justify or verify the use of these proxies itself. This can only be done through empirical, hypothesis-testing approaches with known response observations. What our method can do, however, is to infer, through a weighting resulting from the modelling, which indicators seem to be better suited in terms of their variability and fit with other indicators to trace a population development in the past.

@@

Summing up the probabilities of radiocarbon dates can undoubtedly be considered the standard approach in demographic reconstruction in archaeology today. The logic for this can be derived from the Poisson binomial distribution: such a distribution can be described as the sum of independent Bernoulli trials as it represents the probability estimate for the existence or use of a site per unit time. The expected value per time unit for the number of simultaneously existing sites/settlements is the sum of the individual probabilities for the sites, i.e. the area under the curve of the probabilities of 14C data. Incidentally, the same logic applies to the aoristic sum (shown below). The main problem with this proxy is therefore not so much its statistical nature, but mainly the systematic biases that arise from the conservation and finding practicalities in the research process. To counter these, and to detect and mitigate the systematic biases, hierarchical modelling such as that carried out in this study is appropriate.

@@

We include relative dating information obtained from the heritage authorities of the Swiss cantons (Figure \@ref(fig:aoristmap)). These are primarily derived from scattered surface finds, often with a low dating accuracy (only in the range of archaeological periods), incorporated into our model as a typologically-dated, aorist time series. However, it is not dependent on radiocarbon dating and thus it avoids the methodological issues of sum calibration. Data from `r nrow(aorist)` sites were included in the aoristic sum.

@@

We assume that the higher the population density in an area, the greater the human influence on the natural environment [@zimmermann2004; @lechterbeck2014]. Evidence of deforestation can therefore provide indications of population dynamics. The full procedure for deriving this proxy from several different pollen diagrams is detailed in a previous publication [@heitz2021]. Here, we use five pollen diagrams from sites mainly in the hinterland of the large Alpine lakes.

@@

Our overall model is broken down into several hierarchically-connected individual elements. The process model represents the demographic development itself, without already being explicitly parameterised with data. Here we assume that the latent variable 'number of sites' is strongly autocorrelated across different time periods. The number of sites in 3000 BCE is strongly conditioned by the number of sites in 3050 BCE, and so on. The population at time $t$ results from the population at time $t-1$ times a parameter $\lambda$, which represents the population change at this time.

@@

## Reliability of the overall estimation

We consider each of the proxies used to be intrisically flawed and biased. Therefore, we assume that none of them alone is suitable to generate a population estimate with sufficient credibility. Especially since the use of a single proxy makes it impossible to specify a confidence interval, as its value must be used as a direct indication of a change in population size.

Rather, we assume that all proxies used are influenced, to a greater or lesser extent, by a proportion of random noise as well as by one of the actual signal. As long as there is no ground truth that can be used for calibration, this cannot be fully differentiated. What we assume, however, is that the noise between the individual proxies is uncorrelated, but the signal must essentially produce a coherent pattern.

However, other reasons are also conceivable. For example, a correlation between two parameters can also be caused by the fact that they are affected by the same confounding influence. All archaeologically derived proxies are subject to preservation conditions. If no signs of human activity are preserved for a certain space at a certain time based on the surviving archaeological evidence, then all these indicators will also show a settlement gap. It is therefore all the more important to use independent proxies that are not affected by the same distorting influences for a robust reconstruction. This is the reason why the openness indicator is an essential part of this study, as it is not confounded by the archaeological record. Other similarly independent indicators are to be integrated in future applications of our approach to make the result even more robust.

Moreover, our modelling approach in terms of an abductive reconstructive model is also such that in its present form the model provides the best possible summary of the proxies used and thus the best possible reconstruction given the available information. It must be emphasised that such a model does not attempt to directly represent the process of the relationship between proxy and latent variable. We treat all proxies as equal (false), and leave it to coherence with respect to the underlying process (the latent population development) to provide a weighting. In our view, this corresponds to a (pragmatic) implementation of the general scientific process. However, this also means that we prefer a useful model to a less useful one without wanting to (or being able to) make a statement about the concept of truth of a model.Especially in comparison to the uncritical application of a purely radiocarbon data-based proxy, as it is currently used in large scale, we see the significance of such a model and the philosophy behind it as more than justified.
